Accelerating Neural Network Training using Arbitrary Precision Approximating Matrix Multiplication Algorithms

### 2021-03-07 latex/MPP21/paper.tex
updated abstract with sequential NN speedup number
fix small typos

\begin{abstract}
Matrix multiplication is one of the bottleneck computations for training the weights within deep neural networks.
To speed up the training phase, we propose to use faster algorithms for matrix multiplication known as Arbitrary Precision Approximating (APA) algorithms.
APA algorithms perform asymptotically fewer arithmetic operations than the classical algorithm, but they compute an approximate result with an error that can be made arbitrarily small in exact arithmetic.
Practical APA algorithms provide significant reduction in computation time and still provide enough accuracy for many applications like neural network training.
We demonstrate that APA algorithms can be efficiently implemented and parallelized for shared-memory platforms to obtain up to 28\% and 21\% speedups over the fastest implementation of the classical algorithm using one core and 12 cores, respectively.
Furthermore, using these algorithms to train a Multi-Layer Perceptron (MLP) network yields no significant reduction in the training or testing error.
Our performance results on a large MLP network show overall multithreaded performance improvements of up to 13\%.
\end{abstract}

@@

Overall, we see little fluctuation of the error over matrix dimension, and the theoretical error bound is an upper bound on all empirical errors.
Note that the legend is ordered according to the error parameters, and the empirical error generally follows this ordering.
The two most accurate APA algorithms are the least recently discovered: Bini's $\dims{3,2,2}$ and Schonhage's $\dims{3,3,3;21}$.
Algorithms that offer more potential speedup tend to be less accurate.
Two algorithms with smaller error than expected are $\dims{5,5,5}$ and ${7,2,2}$, which is explained by the coefficients of those algorithms including fraction pre-factors.
While $\phi$ is computed using the exponents of the largest intermediate term of $\lambda$, the leading term of $\dims{5,5,5}$ has a constant of $1/4$, lessening its magnitude. 
We study the effects of the matrix multiplication error on neural network training and test accuracy in \cref{sec:NNacc}.

@@

We adopt the ``hybrid'' parallelization strategy proposed by Benson and Ballard \cite{BB15}.
All linear combinations are parallelized in a straightforward way, in order to maximize the memory bandwidth of the machine.
The multiplications are parallelized as follows: given $r$ multiplications and $p$ threads, with $r=p\cdot q + \ell$ for integers $q$ and $\ell<r$, we assign each thread $q$ multiplications to be performed independently, and the remaining $\ell$ multiplications are performed by all threads using the multithreaded implementation of \texttt{gemm}.
\Cref{fig:hybrid} shows the hybrid strategy for $r=10$ (Bini's algorithm) and $p=4$.

>>>>

\begin{abstract}
Matrix multiplication is one of the bottleneck computations for training the weights within deep neural networks.
To speed up the training phase, we propose to use faster algorithms for matrix multiplication known as Arbitrary Precision Approximating (APA) algorithms.
APA algorithms perform asymptotically fewer arithmetic operations than the classical algorithm, but they compute an approximate result with an error that can be made arbitrarily small in exact arithmetic.
Practical APA algorithms provide significant reduction in computation time and still provide enough accuracy for many applications like neural network training.
We demonstrate that APA algorithms can be efficiently implemented and parallelized for shared-memory platforms to obtain up to 28\% and 21\% speedups over the fastest implementation of the classical algorithm using one core and 12 cores, respectively.
Furthermore, using these algorithms to train a Multi-Layer Perceptron (MLP) network yields no significant reduction in the training or testing error.
Our performance results on a large MLP network show overall sequential and multithreaded performance improvements of up to 25\% and 13\%, respectively.
\end{abstract}

@@

Overall, we see little fluctuation of the error over matrix dimension, and the theoretical error bound is an upper bound on all empirical errors.
Note that the legend is ordered according to the error parameters, and the empirical error generally follows this ordering.
The two most accurate APA algorithms are the least recently discovered: Bini's $\dims{3,2,2}$ and Schonhage's $\dims{3,3,3;21}$.
Algorithms that offer more potential speedup tend to be less accurate.
Two algorithms with smaller error than expected are $\dims{5,5,5}$ and $\dims{7,2,2}$, which is explained by the coefficients of those algorithms including fraction pre-factors.
While $\phi$ is computed using the exponents of the largest intermediate term of $\lambda$, the leading term of $\dims{5,5,5}$ has a constant of $1/4$, lessening its magnitude. 
We study the effects of the matrix multiplication error on neural network training and test accuracy in \cref{sec:NNacc}.

@@

We adopt the ``hybrid'' parallelization strategy proposed by Benson and Ballard \cite{BB15}.
All linear combinations are parallelized in a straightforward way, in order to maximize the memory bandwidth of the machine.
The multiplications are parallelized as follows: given $r$ multiplications and $p$ threads, with $r=p\cdot q + \ell$ for integers $q$ and $\ell<p$, we assign each thread $q$ multiplications to be performed independently, and the remaining $\ell$ multiplications are performed by all threads using the multithreaded implementation of \texttt{gemm}.
\Cref{fig:hybrid} shows the hybrid strategy for $r=10$ (Bini's algorithm) and $p=4$.
